
```{r setup, include=FALSE, messsage=FALSE, warning=FALSE}

library(rmarkdown)
library(tidyverse)
library(lubridate)
library(pollster)


knitr::opts_chunk$set(echo = TRUE)

```

# Household Pulse Survey

The News21 Overview team wanted to see if there was a racial and class disparity in how people experienced the pandemic. We turned to the Household Pulse Survey (HPS), a bi-weekly survey by the U.S. Census Bureau measuring the impact of the coronavirus pandemic from a social and economic perspective. The survey asks questions about childcare, education, employment, food security, health, housing, social security benefits, household spending, consumer spending associated with stimulus payments, intention to receive a COVID-19 vaccination, and transportation.

The HPS produces statistics at the **national** and **state** levels and for the following 15 largest Metropolitan areas:

-   Houston-The Woodlands-Sugar Land, TX Metro Area
-   Riverside-San Bernardino-Ontario, CA Metro Area
-   Los Angeles-Long Beach-Anaheim, CA Metro Area
-   San Francisco-Oakland-Berkeley, CA Metro Area
-   New York-Newark-Jersey City, NY-NJ-PA Metro Area
-   Detroit-Warren-Dearborn, MI Metro Area
-   Dallas-Fort Worth-Arlington, TX Metro Area
-   Miami-Fort Lauderdale-Pompano Beach, FL Metro Area
-   Philadelphia-Camden-Wilmington, PA-NJ-DE-MD Metro Area
-   Chicago-Naperville-Elgin, IL-IN-WI Metro Area
-   Phoenix-Mesa-Chandler, AZ Metro Area

There have been four phases covering the following time periods:

-   Phase 1 (Apr 23, 2020 - July 21, 2020)
-   Phase 2 (Aug 19, 2020 - Oct 26, 2020)
-   Phase 3.1 (Apr 14, 2021 - July 5, 2021)
-   Phase 3.2 (July 21, 2021 - early October)

Phase 1 was collected on a weekly basis, but that proceeding phases had two-week collection periods, eventhough the releases are still called "weekly" releases. Week 1 was also collected over two weeks.

The number of respondents for each collection period varies widely, as does the sample size and response rate. Technical documents, including source and accuracy statements for each release, are included [here](https://www.census.gov/programs-surveys/household-pulse-survey/technical-documentation.html). Overall response rates were low, but, according to a March 2020 [analysis](https://www2.census.gov/programs-surveys/demo/technical-documentation/hhp/2020_HPS_NR_Bias_Report-final.pdf) by the U.S. Census Bureau, weighting adjustment methods have helped mitigate non-response bias, a type of bias that occurs when people are unwilling or unable to respond to a survey due to a factor that makes them differ greatly from people who respond (e.g. internet access).

The U.S. Census Bureau releases the micro-data files containing individual responses to survey questions here: <https://www.census.gov/programs-surveys/household-pulse-survey/datasets.html>

## Analysis

Although we analyzed data from several different points in the pandemic, we settled on Week 22, January 6-18, 2021, when COVID-19 cases peaked in the U.S. The initial health and economic shock of the pandemic caught the country flat footed. By January, however, the U.S. government had been grappling with the pandemic and economic downtown for nine months, and COVID-19 relief under the Trump administration had been widely distributed.

```{r import data, warning=FALSE, Download Data message=FALSE}


download.file("https://www2.census.gov/programs-surveys/demo/datasets/hhp/2021/wk22/HPS_Week22_PUF_SAS.zip", destfile = "./hps22.zip")
unzip("hps22.zip")
hps22 <- read_sas("pulse2021_puf_22.sas7bdat")

```

As our colleague has written, the [poverty rate](https://unmaskingamerica.news21.com/extras/poverty-line/) is an imperfect measurement of the financial health and well-being of a family. The Overview team therefore reviewed responses to 4 questions --- regarding pandemic-related job losses, household expenses, food scarcity, and housing insecurity --- to gauge hardship.

## Re-coding demographics

First, we need to add several columns that re-code variables to get into a format that works for us. We save these changes to a new dataframe, `working_hps`, which we'll be revising throughout the analysis.

The following code chunk re-codes a number of variables: geography, `EST_ST`(State) and `EST_MSA` (Metropolitan statistical area); race and ethnicity, `RHISPANIC` and `RRACE`; gender, `EGENDER`; and education level, `EEDUC`. It also calculates an approximate age from `TBIRTH_year`.Finally, the variable `THHLD_NUMKID` indicates how many people in each household are under the age of 18; the code chunk below uses this to create a flag that indicates whether or not there are children in the household.

```{r re-code demographics, echo=TRUE}

working_hps <- hps22 %>%
  # New state and city variables 
  mutate(state = 
           factor(EST_ST,
                  levels = c('01', '02', '04', '05', '06', '08', '09', 10, 11, 
                               12, 13, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24, 
                               25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 
                               37, 38, 39, 40, 41, 42, 44, 45, 46, 47, 48, 49, 
                               50, 51, 53, 54, 55, 56),
                  # If we don't put quotes around 01-09, they don't code properly. 
                  labels = c("alabama",
                             "alaska",
                             "arizona",
                             "arkansas",
                             "california",
                             "colorado",
                             "connecticut",
                             "delaware",
                             "d.c.",
                             "florida",
                             "georgia",
                             "hawaii",
                             "idaho",
                             "illinois",
                             "indiana",
                             "iowa",
                             "kansas",
                             "kentucky",
                             "louisiana",
                             "maine",
                             "maryland",
                             "massachusetts",
                             "michigan",
                             "minnesota",
                             "mississippi",
                             "missouri",
                             "montana",
                             "nebraska",
                             "nevada",
                             "new hampshire",
                             "new jersey",
                             "new mexico",
                             "new york",
                             "north carolina",
                             "north dakota",
                             "ohio",
                             "oklahoma",
                             "oregon",
                             "pennsylvania",
                             "rhode island",
                             "south carolina",
                             "south dakota",
                             "tennessee",
                             "texas",
                             "utah",
                             "vermont",
                             "virginia",
                             "washington",
                             "west virginia",
                             "wisconsin",
                             "wyoming")
                  ),
          metro =
            factor(EST_MSA,
                   levels = c(35620, 31080, 16980, 19100, 26420, 47900, 33100, 
                              37980, 12060, 38060, 14460, 41860, 40140, 19820, 
                              42660),
                   labels = c("New York-Newark-Jersey City, NY-NJ-PA Metro Area",
                              "Los Angeles-Long Beach-Anaheim, CA Metro Area",
                              "Chicago-Naperville-Elgin, IL-IN-WI Metro Area",
                              "Dallas-Fort Worth-Arlington, TX Metro Area",
                              "Houston-The Woodlands-Sugar Land, TX Metro Area",
                              "Washington-Arlington-Alexandria, DC-VA-MD-WV Metro Area",
                              "Miami-Fort Lauderdale-Pompano Beach, FL Metro Area",
                              "Philadelphia-Camden-Wilmington, PA-NJ-DE-MD Metro Area",
                              "Atlanta-Sandy Springs-Alpharetta, GA Metro Area",
                              "Phoenix-Mesa-Chandler, AZ Metro Area",
                              "Boston-Cambridge-Newton, MA-NH Metro Area",
                              "San Francisco-Oakland-Berkeley, CA Metro Area",
                              "Riverside-San Bernardino-Ontario, CA Metro Area",
                              "Detroit-Warren-Dearborn, MI Metro Area",
                              "Seattle-Tacoma-Bellevue, WA Metro Area")
                    ),
          # New race and ethnicity variable 
          new_race = 
            case_when(RHISPANIC  == 2               ~   10,  # Hispanic
                      RHISPANIC  == 1 & RRACE == 1  ~   20,  # White, non-Hispanic
                      RHISPANIC  == 1 & RRACE == 2  ~   30,  # Black, non-Hispanic
                      RHISPANIC  == 1 & RRACE == 3  ~   40,  # Asian, non-Hispanic
                      RHISPANIC  == 1 & RRACE == 4  ~   50,  # Any other race or combination
                      ),
          race = factor(new_race, 
                        levels = c(10, 20, 30, 40, 50), 
                        labels = c("hispanic", 
                                   "white", 
                                   "black", 
                                   "asian", 
                                   "any other race or combination of races")
                         ),
          # New gender variable
          gender = factor(EGENDER, 
                          levels=c(1,2), 
                          labels=c("male", "female")
                          ),
          # New education
          edu = factor(EEDUC,
                       levels = c(1, 2, 3, 4, 5, 6, 7),
                       labels = c("less than high school",
                                  "some high school",
                                  "high school graduate or equivalent",
                                  "some college",
                                  "associate's degree",
                                  "bachelor's degree",
                                  "graduate degree")
                       ),
          # New income 
          NEW_INCOME =
           case_when(INCOME == 1 | INCOME == 2   ~ 10,  # less than $34,999
                     INCOME == 3 | INCOME == 4   ~ 20,  # $35,000 - $74,999
                     INCOME == 5 | INCOME == 6   ~ 30,  # $75,000 - $149,999
                     INCOME == 7 | INCOME == 8   ~ 40   # $150,000 and above
                     ),
           inc = factor(NEW_INCOME,
                        levels = c(10, 20, 30, 40),
                        labels = c("less than $34,999",
                                   "$35,000 - $74,999",
                                   "$75,000 - $149,999",
                                   "$150,000 and above")
                         ),
          # Children flag
          children = if_else(THHLD_NUMKID > 0, "yes", "no"),
          # Age
          birth_year = lubridate::ymd(TBIRTH_YEAR, truncated = 2L),
          age = age_calc(birth_year, enddate = Sys.Date(), units = "years", precise = FALSE)) %>%
  # Removes the variables we no longer need from df
  select(-TBIRTH_YEAR, -THHLD_NUMKID, -INCOME, -NEW_INCOME, -EEDUC, -EGENDER, -new_race, -RRACE, -RHISPANIC, -EST_MSA, -EST_ST)
```

With the most important demographic information re-coded, we can switch to the responses we care about, starting with household expenses.

## Household Expenses

Q19a asks "In the last 7 days, how difficult has it been for your household to pay for usual household expenses, including but not limited to food, rent or mortgage, car payments, medical expenses, student loans, and so on?" Respondents can select only one answer.

We'll revise `working_hps`, then wrap the new information in its own dataframe, `expns_dif`, for later.

```{r household expenses}

working_hps <- working_hps %>%
  mutate(NEW_EXPNS_DIF = 
            case_when(EXPNS_DIF == 1 | EXPNS_DIF == 2   ~ 10,   # Not difficult or a little difficult
                      EXPNS_DIF == 3 | EXPNS_DIF == 4   ~ 20),  # Somewhat to very difficult
            expns_dif = factor(NEW_EXPNS_DIF,
                               levels = c(10, 20),
                               labels = c("not_difficult", "difficult")
        )) %>%
  select(-NEW_EXPNS_DIF, -EXPNS_DIF)

expns_dif <- working_hps %>%
  select(SCRAM, HWEIGHT, PWEIGHT, expns_dif, state, metro, race, gender, edu, inc, children, birth_year, age)

```

To check our work, we'll create a crosstab that charts percentages of people reporting difficulty paying household expenses across different states -- and then compare those results to [these](https://www.census.gov/data-tools/demo/hhp/#/?measures=EXPENSE&periodSelector=22) from the Census Bureau.

```{r expns_dif by state}

expns_dif %>%
  select(state, expns_dif, PWEIGHT) %>%
  pollster::crosstab(state, expns_dif, weight = PWEIGHT)%>%
  head(5)

```

The results match, which tells us our code is sound. We'll continue to check our results this way for the next two questions regarding food scarcity and housing security.

## Food Insecurity

Q24 asks about household food sufficiency for the last 7 days, with four possible answers: 1) Enough of the kinds of food (I/we) wanted to eat 2) Enough, but not always the kinds of food (I/we) wanted to eat 3) Sometimes not enough to eat 4) Often not enough to eat. Let's condense this into two options: sufficient or insufficient.

```{r}

working_hps <-working_hps %>%
  mutate( NEW_CURFOODSUF = 
            case_when (CURFOODSUF == 1 | CURFOODSUF == 2  ~ 10,   #food sufficient
                       CURFOODSUF == 3 | CURFOODSUF == 4  ~ 30    #food insufficient
                       ),    
          curfoodsuf = 
            factor (NEW_CURFOODSUF,
                    levels = c(10, 30),
                    labels = c("sufficient",
                               "insufficient"
                               ))) %>%
  select(-NEW_CURFOODSUF, -CURFOODSUF)

curfoodsuf <- working_hps %>%
  select(SCRAM, HWEIGHT, PWEIGHT, curfoodsuf, state, metro, race, gender, edu, inc, children, birth_year, age)

```

Again, we'll compare our state results to [those](https://www.census.gov/data-tools/demo/hhp/#/?measures=EXPENSE&periodSelector=22) of the Census Bureau.

```{r}

curfoodsuf %>%
  select(state, curfoodsuf, PWEIGHT) %>%
  pollster::crosstab (state, curfoodsuf, weight = PWEIGHT) %>%
  head(5)

```

Our results for food scarcity match those of the Census Bureau, which verifies our methodology.

## Housing Insecurity

Let's look at housing insecurity across the country next. Several questions address this topic. We'll focus on the following three:

-   Q40b asks whether the household is currently caught up on rent payments.

-   Q40c asks whether the household is currently caught up on mortgage payments.

-   Q41 asks how likely the household will be able to pay the next rent or mortgage payment.

The code chunk below takes responses from all three questions and groups respondents into two buckets. One contains people who are behind on their rent or mortgage payments AND who have little to no confidence they can make their next payments. The other holds everyone else: people who are caught up on rent or mortgage payments, as well as those who've fallen behind but feel confident they can make their next payment. This appears to be how the Census Bureau [measured](https://www.census.gov/data-tools/demo/hhp/#/?measures=HINSEC&periodSelector=22) housing insecurity.

```{r}

working_hps <- working_hps %>%
  mutate(NEW_HOUSING =
           case_when((RENTCUR == 2 | MORTCUR == 2) & (MORTCONF == 1 | MORTCONF == 2)                  ~ 10,     
                     # Behind with little to no confidence
                     RENTCUR == 1 | MORTCUR == 1 | MORTCONF == 3 | MORTCONF == 4 | MORTCONF == 5      ~ 20                                  
                     # Current or confident they can pay
             ),
         housingsec = factor(NEW_HOUSING,
                             levels = c(10, 20),
                             labels = c("housing insecure", "housing secure")
                             )) %>%
  select(-NEW_HOUSING, -RENTCUR, -MORTCUR, -MORTCONF)

housingsec <- working_hps %>%
  select(SCRAM, PWEIGHT, PWEIGHT, housingsec, state, metro, race, gender, edu, inc, children, birth_year, age)

housingsec %>%
  select(state, housingsec, PWEIGHT) %>%
  pollster::crosstab(state, housingsec, weight = PWEIGHT) %>%
  head(5)

```

Our results match [those](https://www.census.gov/data-tools/demo/hhp/#/?measures=HINSEC&periodSelector=22) of the Census Bureau. Good!

## Unemployment Reason

The final variable is from Q13, which asks why a respondent isn't working. There are 12 possible responses. Each respondent can choose only one option. The code chunk below condenses like responses together, as well as accounts for those "currently working," which is answered by Q11, in order to account for the whole population in our analysis.

```{r reason for no work}

working_hps <- working_hps%>%
   mutate (NEW_RSNNOWRK = 
              case_when (RSNNOWRK == 7                                                    ~   10,  
                         # Retired
                         RSNNOWRK == 2  | RSNNOWRK == 3                                   ~   20,  
                         # Sick with coronavirus symptoms or caring for someone who is
                         RSNNOWRK == 13                                                   ~   25,  
                         # Concerned about getting or spreading the coronavirus
                         RSNNOWRK == 4  | RSNNOWRK == 5                                   ~   30,  
                         # Caring for children or an elderly person 
                         RSNNOWRK == 6                                                    ~   40,  
                         # Sick or disabled 
                         RSNNOWRK == 8  | RSNNOWRK == 9 | RSNNOWRK == 10 | RSNNOWRK == 11 ~   50,  
                         # pandemic-related job loss
                         RSNNOWRK == 1  | RSNNOWRK == 12                                  ~   60,  
                         # Other, including those who did not want to be employed at this time 
                         ANYWORK == 1                                                     ~   70   
                         # Currently working
                         ),
             rsnnowrk = factor (NEW_RSNNOWRK, 
                                levels = c(10, 20, 25, 30, 40, 50, 60, 70), 
                                labels = c("retired", 
                                           "sick with coronavirus symptoms or caring for someone who is", 
                                           "concerned about getting or spreading the coronavirus", 
                                           "caring for children or an elderly person", 
                                           "sick or disabled", 
                                           "pandemic-related job loss", 
                                           "other",
                                           "currently working"
                                           ))) %>%
  select(-NEW_RSNNOWRK, -RSNNOWRK, -ANYWORK)


rsnnowrk <- working_hps %>%
  select(SCRAM, PWEIGHT, rsnnowrk, state, metro, race, gender, edu, inc, children, birth_year, age)

```

The U.S. Census Bureau hasn't published a breakdown of this variable by geography, so we can't verify the code like we did for the other variables. But, given that our other calculations match the Census Bureau's, we are confident this code is accurate as well.

Now, let's create a new dataframe that contains the re-coded variables we care about. 

```{r}

recoded_hps <- working_hps %>%
  select(SCRAM, PWEIGHT, rsnnowrk, housingsec, curfoodsuf, expns_dif, state, metro, race, gender, edu, inc, children, birth_year, age)


```

# Analysis

Next, let's see if there's a racial disparity. Although it makes more sense to look at household weights for this question, we're sticking to PWEIGHT to create consistency with other responses. Again, we'll place the results in their own dataframes, so we can't print these tables out later for the data visualization.

```{r expns race}

expns_dif_race <- expns_dif %>%
  select(race, expns_dif, PWEIGHT) %>%
  pollster::crosstab (race, expns_dif, weight = PWEIGHT) %>%
  select(race, "difficult")

expns_dif_race

```

Racial disparities are reproduced in difficulty meeting household expenses. Hispanic and Black individuals were each nearly twice as likely as white households to report having difficulty covering household expenses.

Let's see if there's a class disparity.

```{r expns inc}

expns_dif_inc <- expns_dif %>%
  select(inc, expns_dif, PWEIGHT) %>%
  pollster::crosstab (inc, expns_dif, weight = PWEIGHT) %>%
  select(inc, "difficult")

expns_dif_inc

```

This is rather intuitive: those earning less have harder time meeting household expenses. That said, the disparity is quite stark. Nearly two thirds of people making less than \$35,000 report difficulty; note that \$25,000 is above the poverty line for a family of four.

Food security next

```{r}

curfoodsuf_race <- curfoodsuf %>%
  select(race, curfoodsuf, PWEIGHT) %>%
  pollster::crosstab (race, curfoodsuf, weight=PWEIGHT)%>%
  select(-sufficient, -n)


curfoodsuf_race

```

Class disparity:

```{r}

curfoodsuf_inc <- curfoodsuf %>%
  select(inc, curfoodsuf, PWEIGHT) %>%
  pollster::crosstab (inc, curfoodsuf, weight=PWEIGHT) %>%
  select(-n, -sufficient)

```

A lot of survey takers did not respond to this quesion. Pollster takes these out, so it should no effect on calculations. Let's put `housingcur` into it's own dataframe, and see if there's a racial disparity.

```{r}

housingsec <- working_hps %>%
  select(SCRAM, HWEIGHT, PWEIGHT, housingsec, state, metro, race, gender, edu, inc, children, birth_year, age)

housingsec_race <- housingsec %>%
  select(housingsec, race, PWEIGHT) %>%
  pollster::crosstab (race, housingsec, weight = PWEIGHT) %>%
  select(-n, -"housing secure")

housingsec_inc <- housingsec %>%
  select(housingsec, inc, PWEIGHT) %>%
  pollster::crosstab (inc, housingsec, weight = PWEIGHT) %>%
  select(-n, -"housing secure")

```

Unsurprisingly, the disparity we've tracked is reproduced in housing insecurity, with Hispanic households twice as likely -- and Black households nearly three times as likely -- as white households to be behind on rent or mortgage payments.

With our new variable, we can determine whether there was a racial disparity in pandemic-related job losses, or those who lost work because their employer was forced to scale back due to the pandemic. We'll place each of these analyses in their own dfs as well, that way we can merge the respones into one table when we're done for visualization.

```{r rsnnowrk race}

rsnnowrk_race <- rsnnowrk %>%
  select(age, race, rsnnowrk, PWEIGHT) %>%
  filter(age >= 18) %>%
  pollster::crosstab (race, rsnnowrk, weight=PWEIGHT)%>%
  select(race, "pandemic-related job loss")

rsnnowrk_race

```

According to these results, Hispanic and Black people were about twice as likely as white people to say they lost a job due to the pandemic.

Next, we'll see if there's a class disparity by looking at income.

```{r rsnnowrk inc}

rsnnowrk_inc <- rsnnowrk %>%
  select(age, inc, rsnnowrk, PWEIGHT) %>%
  filter(age >= 18) %>%
  pollster::crosstab (inc, rsnnowrk, weight = PWEIGHT)%>%
  select(inc, "pandemic-related job loss")

rsnnowrk_inc

```

The less in income a person made, the more likely they were to lose their job due to the pandemic. This makes sense given the sectors that were devastated by stay-at-home orders and the fact that higher paying jobs could be done remotely.

# Results

The entire News21 project breaks down individual topic -- such as housing food insecurity, and mental health -- to explore inequality in these areas. We can see from this analysis that when new COVID-19 cases peaked in early January, low-wage workers and people of color were more likely to report hardship.

```{r}

merged_race <- expns_dif_race %>%
  left_join(rsnnowrk_race) %>%
  left_join(curfoodsuf_race) %>%
  left_join(housingsec_race) %>%
  rename("Difficulty paying household expenses" = "difficult",
         "Unemployed due to the pandemic" = "pandemic-related job loss",
         "Food insufficient" = "insufficient",
         "Housing insecure" = "housing insecure")

merged_inc <- expns_dif_inc %>%
  left_join(rsnnowrk_inc) %>%
  left_join(curfoodsuf_inc) %>%
  left_join(housingsec_inc) %>%
  rename("Difficulty paying household expenses" = "difficult",
         "Unemployed due to the pandemic" = "pandemic-related job loss",
         "Food insufficient" = "insufficient",
         "Housing insecure" = "housing insecure")


write.csv(merged_race, "hps22_byrace.csv")
write.csv(merged_inc, "hps22_byinc.csv")
write.csv(recoded_hps, "recoded_hps22.csv")

```
